\documentclass[twoside, letterpaper, twocolumn]{article}

\title{Presenter Notes: On the Dangers of AI Autism Diagnosis: ICDS Scholars Week 2024}
\author{Noah Duggan Erickson\thanks{Department of Computer Science, College of Science and Technology, Western Washington University}}
\date{17 May 2024}

\begin{document}
\maketitle

\section*{Thanks}
Many thanks to the multitude of faculty who have lent a hand in this project.
{\footnotesize
\begin{itemize}
    \item Dr. Kristen Chmielewski (CHSS/RML) for helping to get the project off the ground
    \item Dr. Brian Hutchinson (CS) for his assistance with ML techniques
    \item Heather Cochran M.Ed. (Woodring) for her guidance in the project
\end{itemize}}
\section{Overview}
\begin{itemize}
    \item Getting started
    \item Technical, free to ask questions
    \item Updates since presentation given in HNRS, added visuals \& ML progress
\end{itemize}
\section{What is AIAD?}
\begin{itemize}
    \item Rigor involved in getting an ASD diagnosis
    \item A rapidly growing field in machine learning research
    \item Nexis search for \texttt{AI (AND) Autism (AND) diagno*} gave 96 results: \textbf{20} of them are from \textbf{2021}, \textbf{10} from \textbf{2022}.
    \item Trying\dots \textit{trying\dots} to employ various machine learning techniques in the detection of ASD
\end{itemize}
\section{Problems in General ML Research}
\begin{itemize}
    \item Taking a step back from AIAD for a second.
    \item What are the problem areas in general ML research?
    \item Models are growing faster than technology
    \item Increasing model size means increasing scale of data needed
    \item In the current hot topic of LLMs such as ChatGPT, where does the data come from? How is it filtered? What bias is intrinsic to the dataset? 
    \item No such thing as a perfect filter. Demographics of who edits Wikipedia?
    \item In a \textbf{2020 UW} study, GPT-2 produced toxic responses to \textit{at best} 17\% of innocuous prompts, even after the best ``conditioning'' the study performed.
\end{itemize}
\newpage
\section{Disability Invisibility}
\begin{itemize}
    \item Won't spend too much time here\dots
    \item Treatment of disabled as sub-human, lesser
    \item In much of ML research, gender and race are frequently ``protected characteristics''. Disability often unmarked; \textit{maybe} a footnote.
    \item Conventional AI bias metrics actively blind to ableism; ``fairness'', tying into WW-II masculinity(?)
\end{itemize}
\section{Defining Autism}
\begin{itemize}
    \item How is autism defined in society?
    \item People who I shall not say their names aloud
    \item Indirect inhumanity
    \item AI ethics predicated on personhood; ``viable knowers''
\end{itemize}
\section{Current AIAD Literature}
\begin{itemize}
    \item What are researchers currently doing?
    \item A few main tracks, looking at two of them.
\end{itemize}
\subsection{Description}
\begin{itemize}
    \item MRI
    \begin{itemize}
        \item Most studies use the ABIDE dataset, an aggregation of 3D and 4D MRI brain scans from various institutions.
        \item 78\% accuracy from taking the average of a 2021 review
        \item Variety of techniques, including CNNs and SVMs
    \end{itemize}
    \item Assessments
    \begin{itemize}
        \item Streamlining traditional assessments
        \item Perpetuating the same biases as those assessments
        \item Optimizing feature selection
    \end{itemize}
\end{itemize}
\subsection{CNN}
\begin{itemize}
    \item Tensorflow/Keras implementation of the CNN from that paper
    \item Paper claims X\% accuracy
    \item Violin plot of predictions: Y-axis scale changes, but distributional equivalence remains.
    \item Accuracy plot: Dev set volatile
    \item Likely some sort of \texttt{nilearn} preprocessing I'm missing
\end{itemize}
\subsection{Random Forest}
\begin{itemize}
    \item \texttt{sklearn} implementation of the pipeline from that paper
    \item Pipeline consists of KNN imputation of missing values, minmax scaling, and a random forest classifier
    \item Serious issue with original paper: Included sum of A-scores feature, defeating the purpose.
    \item Heavy selection bias, since existing assessments are used as the gold standard.
    \item Due to internals of Gini importance, age importance is inflated.
    \item Except yellow, colors represent one-hot groups
    \item Except age, all features are binaries
\end{itemize}
\section{Problems in AIAD}
\begin{itemize}
    \item Conclusions, wrap up
\end{itemize}
\end{document}
